---
title: "Beautiful Tables for Models"
---

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}
```

```{r eval=T, echo=F, message = F, warning = F}
library(tidyverse)
library(gtsummary)
library(medicaldata)
```

## Recap of `ggplot2`

The package `ggplot2` (download as part of the `tidyverse` aggregate package) is the primary publication-suitable plotting package in R. While there are many other domain-specific packages, as well as numerous gg-themed add-on packages, most plots rely on a similar scaffolding plotting logic.

We will start by making some basic summary plots of the `smartpill` data to briefly recap `ggplot2` syntax.

```{r}
head(smartpill)
```

Begin by using the `ggplot()` function to specify input data. Within, use `aes()` to specify variables (columns) to use for x and y and any variable-driven formatting like color. Use the appropriate `geom_()` option to specify type of plot (e.g., scatter plot, bar plot).
```{r label='basic_plot'}
ggplot(data = smartpill, aes(x = SB.Time, y = GE.Time, color = Weight)) +
  geom_point()
```

