---
title: "Customize Tables"
---

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}
```

## Overview of `gt`

With `gt` you can construct a wide variety of useful tables with a cohesive set of table parts. Here is the main workflow to transform a data frame into a publication-ready table:

![https://gt.rstudio.com/reference/figures/gt_workflow_diagram.svg](https://gt.rstudio.com/reference/figures/gt_workflow_diagram.svg)

An `gt` object as various elements. These include the table **header**, the **stub**, the **column labels** and **spanner column labels**, the **table body**, and the **table footer**.

![https://gt.rstudio.com/reference/figures/gt_parts_of_a_table.svg](https://gt.rstudio.com/reference/figures/gt_parts_of_a_table.svg)

# Our example 

We are going to reproduce some of the tables of this paper [Use of wireless motility capsule to determine gastric emptying and small intestinal transit times in critically ill trauma patients](https://doi.org/10.1016/j.jcrc.2011.12.002) and create new ones using the study data in the `smartpill` dataset.


## Load the libraries

```{r label='install_load'}
#| message: FALSE
 
# install.packages("gt")
library(gt)
library(tidyverse)
library(medicaldata)
```

```{r label='group_factor', echo=FALSE}
smartpill <- smartpill %>% 
  mutate(Group = as.factor(Group))
```

## A basic table

Now let's create a summary table of demographics of study participants.

```{r}
#| message: FALSE

table1 <- smartpill  %>% 
  select(Group, Gender, Height, Weight, Age) %>%
  group_by(Group) %>%
    summarise(average_age = mean(Age),
              average_height = mean(Height),
              average_weight = mean(Weight))

table1
```

We can turn this into a `gt` table with function `gt()`:

```{r}
table1_pub <-
  table1 %>%
  #make gt object
  gt() 

table1_pub
```
#### _NOTE TO SELF - transpose example_




## Title
```{r}
#| message: FALSE

table1_pub %>% 
  tab_header(
    title = "Table 1. Demographic and morphometric characteristics",
    subtitle = "smartpill dataset"
  ) 
```

## Column names

## Footnote


```{r}

```


## Styling


```{r}

```


There are 30+ combinations of [style and color](https://gt.rstudio.com/reference/opt_stylize.html?q=opt_stylize#examples) to choose from as well as [conditional cell formatting](https://gt.rstudio.com/reference/tab_style.html).

## Additional resources

* [`gt` website](https://gt.rstudio.com/articles/gt.html)
