---
title: "Customize tables with `gt`"
---

## A few words on the `gt` package


With `gt` you can construct a wide variety of useful tables with a cohesive set of table parts. Here is the main workflow to transform a data frame into a beautiful table:

![https://gt.rstudio.com/reference/figures/gt_workflow_diagram.svg](https://gt.rstudio.com/reference/figures/gt_workflow_diagram.svg)

An `gt` object as various elements. These include the table **header**, the **stub**, the **column labels** and **spanner column labels**, the **table body**, and the **table footer**.

![https://gt.rstudio.com/reference/figures/gt_parts_of_a_table.svg](https://gt.rstudio.com/reference/figures/gt_parts_of_a_table.svg)

Checkout the `gt` website for more information: <https://gt.rstudio.com/articles/gt.html>

# Our example 

We are going to reproduce some of the tables of this paper [Use of wireless motility capsule to determine gastric emptying and small intestinal transit times in critically ill trauma patients](https://doi.org/10.1016/j.jcrc.2011.12.002) and create new ones using the study data in the `smartpill` dataset.


## Load the libraries

```{r label='install_load'}
#| message: FALSE
 
# install.packages("gt")
library(gt)
library(tidyverse)
library(medicaldata)
```

## Load the data

```{r label='read_in_data'}
#included in the medicaldata package
attach(smartpill)
data(smartpill)
```


## Have a look at the data

```{r label='summary'}
smartpill |> glimpse()
```


Let's have a look at the first rows of our dataset, but this time we are going to use the `gt` package

```{r}
smartpill |> 
  head() |>
  gt()
```

## A simple table

Now let's create a summary table of demographics of study participants.

```{r}
#| message: FALSE

table1 <- smartpill |>
  select(Group, Gender, Height, Weight, Age) |>
  group_by(Group) |>
    summarise(average_age = mean(Age),
              average_height = mean(Height),
              average_weight = mean(Weight))

table1
```


## Now let's make this a nice table

```{r}
table1_pub <-
  table1 |>
  gt() 

# Show the gt Table
table1_pub
```
#### _NOTE TO SELF - transpose example_




## Add a Title
```{r}
#| message: FALSE

table1_pub |> 
  tab_header(
    title = "Table 1. Demographic and morphometric characteristics",
    subtitle = "smartpill dataset"
  ) 
```



## Add better column names

```{r}
gt_measurements <- data_measurements |>
  gt() |>
  cols_label(
    avg_air_temp = "Air Temp (C)",
    avg_pco2_atmosphere_ppm = "Air pC02 (ppm)",
    avg_h2o_temp_c = "Water Temp (C)",
    avg_ph = 'Water pH',
    avg_pco2_water_ppm = "Water pC02 (ppm)"
  )


# Show the gt Table
gt_measurements
```


## Adding headers for multiple columns

```{r}
gt_measurements <- gt_measurements |>
  tab_spanner(
    label = md("**Air**"),
    columns = c(avg_air_temp, avg_pco2_atmosphere_ppm)
  ) |>
  tab_spanner(
    label = md("**Water**"),
    columns = c(avg_h2o_temp_c, avg_ph, avg_pco2_water_ppm)
  ) |>
  cols_label(
    avg_air_temp = "Temp (C)",
    avg_pco2_atmosphere_ppm = "pC02 (ppm)",
    avg_h2o_temp_c = "Temp (C)",
    avg_ph = 'pH',
    avg_pco2_water_ppm = "pC02 (ppm)"
  )

# Show the gt Table
gt_measurements
```



## Limiting values to 2 decimals

```{r}
gt_measurements <- gt_measurements |>
    fmt_number(
    columns = everything(),
    decimals = 2,
    use_seps = FALSE
  )


# Show the gt Table
gt_measurements
```

## Put it all together

### Add title and Footnote


```{r}
gt_measurements <- gt_measurements |>
  tab_header(
    title = "Average measurements per lake and year",
    subtitle = "major reservoirs in San Diego County"
  ) |>
 tab_source_note(
    source_note = md("Source: _Adamczyk EM, Shurin JB (2015) Seasonal Changes in Plankton Food Web Structure and Carbon Dioxide Flux from Southern California Reservoirs. PLoS ONE 10(10): e0140464. <https://doi.org/10.1371/journal.pone.0140464>_")
  )

# Show the gt Table
gt_measurements
```


## Styling


```{r}
gt_measurements |> opt_stylize(style = 6, color = 'gray')
```


There are 36 combinations of style and color to choose from: <https://gt.rstudio.com/reference/opt_stylize.html?q=opt_stylize#examples>

And of course you can do conditional cell formatting: <https://gt.rstudio.com/reference/tab_style.html>
