---
title: "Create Statistical Tables"
---

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}
```

```{r eval=T, echo=F, message = F, warning = F}
library(tidyverse)
library(gt)
library(gtsummary)
library(medicaldata)
```

```{r label='group_factor', echo=FALSE}
smartpill <- smartpill %>% 
  mutate(Group = as.factor(Group))
```

## Introduction to `gtsummary`

Next, we'll use the `gtsummary` package, an extension of the `gt` table formatting package, that can incorporate statistical model outputs into publication-ready tables.

## Descriptive statistics

The base R approach to getting summary statistics for a dataframe results in a thorough, but not well-formatted for publication, table.

```{r label='base_summary'}
summary(smartpill)
```

The equivalent summary statistics function from `gtsummary` is `tbl_summary()`.

```{r label='tbl_summary'}
tbl_summary(smartpill) 
```


### Change default descriptive values

The default summary statistic presented depends on variable type. Categorical variables get counts and percentages while continuous variables get `median(first quartile, third quartile)`. Any NA values are listed in a separate row as "unknown".

These defaults are customizable. For instance, if we wanted only a subset of variables:

```{r label='customize_descriptive'}
smartpill %>%  
  tbl_summary(
    include = c(Group, Age, SB.Time, WG.Time)
  )
```

The default summary statistic presented depends on variable type. Categorical variables are displayed as counts and percentages while continuous variables are displayed as `median(first quartile, third quartile)`. Any NA values are listed in a separate row as "unknown".

These defaults are customizable. For instance, if we wanted to display continuous variables as minimum, median, and maximum:

```{r label='customize_continuous'}
smartpill %>%  
  tbl_summary(
    include = c(Group, Age, SB.Time, WG.Time), 
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    )
  )
```

### Variable labels and NAs

This table can also be customized with better labels for rows (variables) and updated NA indicator:

```{r label='customize_labels'}
smartpill %>%  
  tbl_summary(
    include = c(Group, Age, SB.Time, WG.Time), 
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    ),
    #update row labels
    label = list(Group = "Patient Group",
                 SB.Time = "Small bowel transit time (hours)", 
                 WG.Time = "Whole gut time (hours)"),
    #change text for NA values
    missing_text = "Missing"
  )
```

Note that if we want to change any categorical/character variable _values_, this needs to be done outside `tbl_summary`. Example process to reset factor level names of Group variable:

```{r}
#check existing levels
levels(smartpill$Group)
```

```{r}
#count of each level
summary(smartpill$Group)
```

```{r}
#relace level values
levels(smartpill$Group) <- c("Critically Ill", "Healthy")
```

```{r}
#confirm accurate counts
summary(smartpill$Group)
```

```{r}
smartpill %>%  
  tbl_summary(
    include = c(Group, Age, SB.Time, WG.Time), 
    statistic = list(
      all_continuous() ~ "{median} ({min}, {max})"
    ),
    label = list(Group = "Patient Group",
                 SB.Time = "Small bowel transit time (hours)", 
                 WG.Time = "Whole gut time (hours)"),
    missing_text = "Missing"
  )
```

### Crosstabs and column names

This structure can be extended to produce crosstab tables, for instance with descriptive statistics for selected variables for each group

```{r label='crosstab'}
smartpill %>%  
  tbl_summary(
    by = Group,
    include = c(Age, SB.Time, WG.Time), 
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    ),
    label = list(SB.Time = "Small bowel transit time (hours)", 
                 WG.Time = "Whole gut time (hours)"),
    missing_text = "Missing"
  )
```

_Note_: Warnings will be returned if the metrics are not workable, for instance if all instances in a group are not missing. If we created this table with variable 'SB.Contractions', we would see NaN values and -Inf as outputs:

```{r label='error'}
smartpill %>%  
  tbl_summary(
    by = Group,
    include = c(Age, SB.Time, WG.Time), 
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    ),
    label = list(SB.Time = "Small bowel transit time (hours)", 
                 WG.Time = "Whole gut time (hours)"),
    missing_text = "Missing"
  )
```

If needed, [column names can be updated](https://www.danieldsjoberg.com/gtsummary/reference/modify.html).

```{r label='col_names'}
smartpill %>%  
  mutate(Group = as.factor(Group)) %>%
  tbl_summary(
    by = Group,
    include = c(Age, SB.Time, WG.Time), 
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    ),
    label = list(SB.Time = "Small bowel transit time (hours)", 
                 WG.Time = "Whole gut time (hours)"),
    missing_text = "Missing"
  ) %>% 
  #show existing column headers of this table
  show_header_names()

```

```{r label='modify_cols'}
smartpill %>%  
  tbl_summary(
    by = Group,
    include = c(Age, SB.Time, WG.Time), 
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    ),
    label = list(SB.Time = "Small bowel transit time (hours)", 
                 WG.Time = "Whole gut time (hours)"),
    missing_text = "Missing"
  ) %>% 
  #modify column header text and make bold with **
  modify_header(label = "**Variable**", 
                stat_1 = "**Critically Ill Trauma Patients**", 
                stat_2 = "**Healthy Volunteers**")
```

Before we move on to the next section, let's assign our current table to an object to make it easier downstream to see what code is being added.

```{r label='table1'}
Table1 <- smartpill %>%  
  tbl_summary(
    by = Group,
    include = c(Age, SB.Time, WG.Time), 
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    ),
    label = list(SB.Time = "Small bowel transit time (hours)", 
                 WG.Time = "Whole gut time (hours)"),
    missing_text = "Missing"
  ) %>% 
  #modify column header text and make bold with **
  modify_header(label = "**Variable**", 
                stat_1 = "**Critically Ill Trauma Patients**", 
                stat_2 = "**Healthy Volunteers**")
```

## Statistical tests

If we wanted to run statistical analyses on this data, a usual approach would be to use a separate R package to run the test and then extract values of interest, like coefficients and p-values, to put into a table. `gtsummary` streamlines this process with an all-in-one approach to formatting statistical test outputs into a table.

There are many standard [statistical tests integrated into gtsummary](https://www.danieldsjoberg.com/gtsummary/reference/tests.html) including t-test, ANOVA, chi-square, regression, survey sample methods, and more.

Let's look at a couple examples in practice.

Comparing small bowel transit time and whole gut time bewteen the two groups:

```{r label='stat_compare'}
Table1 %>% 
  #add p-value
  add_p()
```

### Caveats

Default tests will vary based on data type.

Since we did not define a specific test to use, `add_p()` used the [default tests](https://www.danieldsjoberg.com/gtsummary/reference/add_p.tbl_summary.html), which depend on whether the data is continuous or categorical, how many categories, etc. The test used is always noted at the bottom of the table.

>The default test used in add_p() primarily depends on these factors: 
>
>* whether the variable is categorical/dichotomous vs continuous 
>* number of levels in the tbl_summary(by) variable
>* whether the add_p(group) argument is specified
>* whether the add_p(adj.vars) argument is specified

In this case, the variable being compared was continuous and the grouping variable had two levels, so Wilcoxon rank sum test was used.

If for any reason you want to override the default test, you can specify a different test within the `add_p()` function.

```{r label='stat_compare_2'}
Table1 %>% 
  #add p-value, override default to specify which test to use for Age (here, a t-test)
  add_p(test = Age ~ "t.test")
```


## Further customization 

For some customizations, we'll need to convert our `gtsummary` object to a `gt` object with the [`as_gt()` function](https://www.danieldsjoberg.com/gtsummary/reference/as_gt.html). 

![https://gt.rstudio.com/reference/figures/gt_parts_of_a_table.svg](https://gt.rstudio.com/reference/figures/gt_parts_of_a_table.svg)

For instance, to update the footnote to clarify which test was used on which variable.

```{r label='stat_compare_3'}
Table1 %>% 
  add_p(test = Age ~ "t.test") %>% 
  #convert to gt object keeping everything except the existing footnotes
  as_gt(include = -tab_footnote) %>% 
  #add new footnote specifying which test used for which variable
  tab_footnote(footnote = "t.test for Age; Wilcox rank sum for others",
               locations = cells_column_labels(columns = c(p.value)))
```

Or to add a title and/or subtitle to the table:

```{r label='table_title'}
Table1 %>% 
  add_p(test = Age ~ "t.test") %>% 
  as_gt(include = -tab_footnote) %>% 
  tab_footnote(footnote = "t.test for Age; Wilcox rank sum for others",
               locations = cells_column_labels(columns = c(p.value))) %>% 
  #add table title and subtitle
  tab_header(title = "Table 1",
             subtitle = "smartpill dataset")
```

As with plots, there are many options for table styling: [36 combinations of style and color to choose from](https://gt.rstudio.com/reference/opt_stylize.html?q=opt_stylize#examples) plus [conditional cell formatting](https://gt.rstudio.com/reference/tab_style.html)

```{r label='styling'}
Table1 %>% 
  add_p(test = Age ~ "t.test") %>% 
  as_gt(include = -tab_footnote) %>% 
  tab_footnote(footnote = "t.test for Age; Wilcox rank sum for others",
               locations = cells_column_labels(columns = c(p.value))) %>% 
  tab_header(title = "Table 1",
             subtitle = "smartpill dataset") %>% 
  #styling function
  opt_stylize(style = 6, color = 'gray')
```

## Saving tables

The [function `gtsave()`](https://gt.rstudio.com/reference/gtsave.html) is the parallel table save function to `ggsave()` for plots.

```{r label='saving'}
#assign updates to Table1 object
Table1_save <- Table1 %>% 
          add_p(test = Age ~ "t.test") %>% 
          as_gt(include = -tab_footnote) %>% 
          tab_footnote(footnote = "t.test for Age; Wilcox rank sum for others",
                       locations = cells_column_labels(columns = c(p.value))) %>% 
          tab_header(title = "Table 1",
                     subtitle = "smartpill dataset") %>% 
          opt_stylize(style = 6, color = 'gray')
```

Specify object, filename, and path (if necessary, defaults to current working directory). Options for output format are .html, .tex, .ltx, .rtf, .docx.

Example saving as Word file:
```{r warning=FALSE, eval=FALSE}
gtsave(data = Table1_save, filename = "Table1.docx")
```

_Caveat_: Word files sometimes struggle to retain proper color formatting. 

Saving a PNG is possible - it results in a cropped image of an HTML table. The amount of whitespace can be set with the expand option.
```{r warning=FALSE, eval=FALSE}
gtsave(data = Table1_save, filename = "Table1.png", expand = 5)
```

__Note__: You may get a warning 'The package "webshot2" is required to save gt tables as images.' R will prompt you to install `webshot2` if you want to save images of tables like this.

## Additional resources

* [Primary website for gtsummary](https://www.danieldsjoberg.com/gtsummary/) including how to [cite use of the package](https://www.danieldsjoberg.com/gtsummary/authors.html#citation).
* gtsummary [reference manual](https://cran.r-project.org/web/packages/gtsummary/refman/gtsummary.html) with complete details on functions and arguments.
* Examples of [gtsummary in practice from R Graph Gallery](https://r-graph-gallery.com/package/gtsummary.html). 
* [Primary website for `gt`](https://gt.rstudio.com/articles/gt.html)
