---
title: "Beautiful Tables for Models"
---

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}
```

## Introduction to `gtsummary`

Next, we'll use the `gtsummary` package, an extension of `gt` that can incorporate statistical model outputs into publication-ready tables.

## Descriptive statistics

The base R approach to getting summary statistics for a dataframe results in a thorough, but not well-formatted for publication, table.

```{r label='base_summary'}
summary(smartpill)
```

The equivalent summary statistics function from `gtsummary` is `tbl_summary()`.

```{r label='tbl_summary'}
tbl_summary(smartpill)
```


The default summary statistic presented depends on variable type. Categorical variables get counts and percentages while continuous variables get `median(first quartile, third quartile)`. Any NA values are listed in a separate row as "unknown".

These defaults are customizable. For instance, if we wanted only a subset of variables:

```{r label='customize_descriptive'}
smartpill |>  
  tbl_summary(
    include = c(Group, Age, SB.Time, SB.Contractions)
  )
```

We can incorporate other R functions before using `tbl_summary()` to convert any unintentionally numeric variables to categories so they are displayed in the table as distinct groups.

```{r label='customize_factors'}
smartpill |>  
  mutate(Group = as.factor(Group)) |>
  tbl_summary(
    include = c(Group, Age, SB.Time, SB.Contractions)
  )
```

The default summary statistic presented depends on variable type. Categorical variables get counts and percentages while continuous variables get `median(first quartile, third quartile)`. Any NA values are listed in a separate row as "unknown".

These defaults are customizable. For instance, if we wanted to display continuous variables as minimum, median, and maximum:

```{r label='customize_continuous'}
smartpill |>  
  mutate(Group = as.factor(Group)) |>
  tbl_summary(
    include = c(Group, Age, SB.Time, SB.Contractions), 
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    )
  )
```

This table can also be customized with better labels and updated NA indicator:

```{r label='customize_labels'}
smartpill |>  
  mutate(Group = as.factor(Group)) |>
  tbl_summary(
    include = c(Group, Age, SB.Time, SB.Contractions), 
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    ),
    label = list(SB.Time = "Small bowel transit time (hours)", SB.Contractions = "Small bowel contractions"),
    missing_text = "Missing"
  )
```

Finally, this can be extended to produce crosstab tables, for instance with descriptive statistics for selected variables for each group

```{r label='crosstab'}
smartpill |>  
  mutate(Group = as.factor(Group)) |>
  tbl_summary(
    by = Group,
    include = c(Age, SB.Time, SB.Contractions), 
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    ),
    label = list(SB.Time = "Small bowel transit time (hours)", SB.Contractions = "Small bowel contractions"),
    missing_text = "Missing"
  )
```

Note that warnings will be returned if the metrics are not workable, for instance if all instances in a group are not missing.


## Statistical tests

If we wanted to run statistical analyses on this data, a usual approach would be to use a separate R package to run the test and then extract values of interest, like coefficients and p-values, to put into a table. `gtsummary` streamlines this process with an all-in-one approach to formatting statistical test outputs into a table.

There are many standard [statistical tests integrated into gtsummary](https://www.danieldsjoberg.com/gtsummary/reference/tests.html) including t-test, ANOVA, chi-square, regression, survey sample methods, and more.

Let's look at a couple examples in practice.





### Table with significance

Run a one-way ANOVA to compare whether chlorophyll is significantly different between the three lakes in 2013:

```{r}
oneway.test(chlorophyll_a_ugL ~ as.factor(lake), data = data)
```


If we wanted to report these results, we'd have some work to do to paste this into a table format. Or, we can do the same analysis using the [`tbl_summary`](https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html) function from `gtsummary` to do this for us!

> The tbl_summary() function calculates descriptive statistics for continuous, categorical, and dichotomous variables in R, and presents the results in a beautiful, customizable summary table ready for publication (for example, Table 1 or demographic tables).

Let's run the same one-way ANOVA as above:

```{r}
data |> 
  #specify what variables we want in our table: here, we want a row for chlorophyll with columns for lakes
  tbl_summary(by = lake, include = c(chlorophyll_a_ugL)) |>
  #specify what test to perform on which variables, to get p-values from
  #here, a one-way ANOVA on chlorphyll
  add_p(test = chlorophyll_a_ugL ~ "oneway.test")
```


#### Caveats

**1)** Structure `tbl_summary()` in line with which test will be run in `add_p()`.

Note that if we don't include table `by = lake` in `tbl_summary()`, the test won't know which groups to compare for this ANOVA.

```{r}
#| error: true
data |> 
  #no specification that we are grouping our data into lake-specific columns
  tbl_summary(include = c(chlorophyll_a_ugL)) |>
  add_p(test = chlorophyll_a_ugL ~ "oneway.test")
```

**2)** Default tests will vary based on data type.

We don't specify a test, `add_p()` will use the [default tests](https://www.danieldsjoberg.com/gtsummary/reference/add_p.tbl_summary.html), which depend on whether the data is continuous or categorical, how many categories, etc.

If we don't specify anova, the analysis will default to a Kruskal-Wallis rank sum test.

>The default test used in add_p() primarily depends on these factors: 
>
>* whether the variable is categorical/dichotomous vs continuous 
>* number of levels in the tbl_summary(by) variable
>* whether the add_p(group) argument is specified
>* whether the add_p(adj.vars) argument is specified

```{r}
data |> 
  tbl_summary(by= lake, include = c(chlorophyll_a_ugL)) |>
  add_p()
```


### Table with regression results

When running a regression, we would want to have all model results - coefficients, confidence interval, and p-value - formatted into a table.

Predict chlorophyll from water temperature, nitrogen, phosphorus, and dissolved organic carbon:

```{r}
#use lm() function to create model object
mod1 <- lm(chlorophyll_a_ugL ~ h2o_temp_c + 
                              total_nitrogen_mgL + 
                              total_phosphorus._mgL + 
                              dissolved_organic_carbon_mgL, 
           data = data)
```

```{r}
#use summary() to view result
summary(mod1)
```


The equivalent function in `gtsummary` is `tbl_regression()`, which produces a formatted table of regression results.

```{r}
tbl_regression(mod1)
```

This table is customizable, using a [mix of `gt` and `gtsummary` functions for formatting](https://www.danieldsjoberg.com/gtsummary/articles/tbl_regression.html#customize-output) the output table.

```{r}
tbl_regression(mod1,
               #update variable names
               label = list(h2o_temp_c = "Water Temperature (C)", 
                            total_nitrogen_mgL = "Total Nitrogen (mgL)", 
                            total_phosphorus._mgL = "Total Phosphorus (mgL)", 
                            dissolved_organic_carbon_mgL = "Dissolved Organic Carbon (mgL)")) |>
  #make gt object, in order to use gt functions below
   as_gt() |>
  #add source note
   gt::tab_source_note(source_note = md("Source: _Adamczyk EM, Shurin JB (2015) Seasonal Changes in Plankton Food Web Structure and Carbon Dioxide Flux from Southern California Reservoirs. PLoS ONE 10(10): e0140464. https://doi.org/10.1371/journal.pone.0140464_")
  )
```


## Additional resources

* [Primary website for gtsummary](https://www.danieldsjoberg.com/gtsummary/) including how to [cite use of the package](https://www.danieldsjoberg.com/gtsummary/authors.html#citation).
* gtsummary [reference manual](https://cran.r-project.org/web/packages/gtsummary/refman/gtsummary.html) with complete details on functions and arguments.
* Examples of [gtsummary in practice from R Graph Gallery](https://r-graph-gallery.com/package/gtsummary.html). 
