---
title: "Create Statistical Tables"
---

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}
```

```{r eval=T, echo=F, message = F, warning = F}
library(tidyverse)
library(gtsummary)
library(medicaldata)
```


## Introduction to `gtsummary`

Next, we'll use the `gtsummary` package, an extension of `gt` that can incorporate statistical model outputs into publication-ready tables.

## Descriptive statistics

The base R approach to getting summary statistics for a dataframe results in a thorough, but not well-formatted for publication, table.

```{r label='base_summary'}
summary(smartpill)
```

The equivalent summary statistics function from `gtsummary` is `tbl_summary()`.

```{r label='tbl_summary'}
tbl_summary(smartpill)
```


The default summary statistic presented depends on variable type. Categorical variables get counts and percentages while continuous variables get `median(first quartile, third quartile)`. Any NA values are listed in a separate row as "unknown".

These defaults are customizable. For instance, if we wanted only a subset of variables:

```{r label='customize_descriptive'}
smartpill |>  
  tbl_summary(
    include = c(Group, Age, SB.Time, SB.Contractions)
  )
```

We can incorporate other R functions before using `tbl_summary()` to convert any unintentionally numeric variables to categories so they are displayed in the table as distinct groups.

```{r label='customize_factors'}
smartpill |>  
  mutate(Group = as.factor(Group)) |>
  tbl_summary(
    include = c(Group, Age, SB.Time, SB.Contractions)
  )
```

The default summary statistic presented depends on variable type. Categorical variables get counts and percentages while continuous variables get `median(first quartile, third quartile)`. Any NA values are listed in a separate row as "unknown".

These defaults are customizable. For instance, if we wanted to display continuous variables as minimum, median, and maximum:

```{r label='customize_continuous'}
smartpill |>  
  mutate(Group = as.factor(Group)) |>
  tbl_summary(
    include = c(Group, Age, SB.Time, SB.Contractions), 
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    )
  )
```

This table can also be customized with better labels and updated NA indicator:

```{r label='customize_labels'}
smartpill |>  
  mutate(Group = as.factor(Group)) |>
  tbl_summary(
    include = c(Group, Age, SB.Time, SB.Contractions), 
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    ),
    label = list(SB.Time = "Small bowel transit time (hours)", SB.Contractions = "Small bowel contractions"),
    missing_text = "Missing"
  )
```

Finally, this can be extended to produce crosstab tables, for instance with descriptive statistics for selected variables for each group

```{r label='crosstab'}
smartpill |>  
  mutate(Group = as.factor(Group)) |>
  tbl_summary(
    by = Group,
    include = c(Age, SB.Time, SB.Contractions), 
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    ),
    label = list(SB.Time = "Small bowel transit time (hours)", SB.Contractions = "Small bowel contractions"),
    missing_text = "Missing"
  )
```

Note that warnings will be returned if the metrics are not workable, for instance if all instances in a group are not missing.


## Statistical tests

If we wanted to run statistical analyses on this data, a usual approach would be to use a separate R package to run the test and then extract values of interest, like coefficients and p-values, to put into a table. `gtsummary` streamlines this process with an all-in-one approach to formatting statistical test outputs into a table.

There are many standard [statistical tests integrated into gtsummary](https://www.danieldsjoberg.com/gtsummary/reference/tests.html) including t-test, ANOVA, chi-square, regression, survey sample methods, and more.

Let's look at a couple examples in practice.





## Additional resources

* [Primary website for gtsummary](https://www.danieldsjoberg.com/gtsummary/) including how to [cite use of the package](https://www.danieldsjoberg.com/gtsummary/authors.html#citation).
* gtsummary [reference manual](https://cran.r-project.org/web/packages/gtsummary/refman/gtsummary.html) with complete details on functions and arguments.
* Examples of [gtsummary in practice from R Graph Gallery](https://r-graph-gallery.com/package/gtsummary.html). 
