---
title: "Demo: Sequencing Plots"
---

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}
```

```{r eval=T, echo=F, message = F, warning = F}
library(tidyverse)
library(medicaldata)
```

## ggplot extensions for specific data types

Along with statistical annotation extensions of `ggplot2`, there are also [many extension packages](https://exts.ggplot2.tidyverse.org/) designed to visualize specific types or formats of data, like sequencing data files or genomics data formats.

This section includes a brief demo of three of these packages:
* [`ggDNAvis`](https://ejade42.github.io/ggDNAvis/)
* [`gggenomes`](https://thackl.github.io/gggenomes/)
* [`ggtranscript`](https://dzhang32.github.io/ggtranscript/)


### Sequence strings
The [`ggDNAviz`](https://ejade42.github.io/ggDNAvis/index.html) package can make ggplot-style plots for genetic data:
| a single DNA/RNA sequence split across multiple lines,
| multiple DNA/RNA sequences, each occupying a whole line, or
| base modifications such as DNA methylation called by modified-bases models in Dorado or Guppy.

```{r eval=FALSE}
#Install package (only the first time)
install.packages("ggDNAvis")
```

```{r}
library(ggDNAvis)
library(tidyverse) #loads ggplot2 and other useful dependencies
```

Let's recreate some of the plots in the official pcakge documentation, using the built-in dataset `example_many_sequences`.

First, check out the sample data:

```{r}
str(example_many_sequences)
```

```{r}
head(example_many_sequences)
```

To create a ggplot-style visualization of sequence data:

First, use one of the [data processing function](https://ejade42.github.io/ggDNAvis/reference/index.html#key-data-processing-functions) to extract and sort the sequence data from the full dataframe.

```{r}
sequences <- extract_and_sort_sequences(example_many_sequences)

head(sequences)
```
Then, use the [`visualize_many_sequqneces()` function](https://ejade42.github.io/ggDNAvis/reference/visualise_many_sequences.html). We will save the image to a file within the same function, then import it back in to view. (If following along locally, you can directly open the created png from your local file explorer to view.)

```{r}
#viewing in R isn't great; write to file image
visualize_many_sequences(sequences_vector = sequences, 
                         filename = "ggDNAviz_plot.png",
                         #don't also show image in R
                         return = FALSE)
```

```{r}
#view created image
image <- png::readPNG("ggDNAviz_plot.png")
unlink("ggDNAviz_plot.png")
grid::grid.newpage()
grid::grid.raster(image)
```

Colors for ACTG are the default ggplot-style colors, but can be updating using the [sequence_colors](https://ejade42.github.io/ggDNAvis/reference/visualise_many_sequences.html#arg-sequence-colours) argument within the `visualize_many_sequences()` function.

Next, update this image to visualize methylation probabilities for these sequences.

Use the methylation [data processing function](https://ejade42.github.io/ggDNAvis/reference/index.html#key-data-processing-functions) to extract relevant data.

```{r}
methylation_info <- extract_and_sort_methylation(example_many_sequences)

str(methylation_info)
```
Then use the [`visualise_methylation()` function](https://ejade42.github.io/ggDNAvis/reference/visualise_methylation.html) to create a new image with this information. As before, save the image within the function and then read in to view. 

There are many customization options available:

```{r}
visualise_methylation(
  #specify where to get data for locations, probabilities, and sequences
    modification_locations = methylation_info$locations,
    modification_probabilities = methylation_info$probabilities,
    sequences = methylation_info$sequences,
  #set colors and annotations
    low_colour = "white",
    high_colour = "black",
    index_annotation_lines = c(1, 23, 37),
    index_annotation_interval = 3,
    index_annotation_full_line = FALSE,
    index_annotation_always_first_base = TRUE,
    index_annotation_always_last_base = TRUE,
    other_bases_colour = "lightblue1",
    other_bases_outline_linewidth = 1,
    other_bases_outline_colour = "grey",
    modified_bases_outline_linewidth = 3,
    modified_bases_outline_colour = "black",
    margin = 0.3,
  #save image to png file, do not also show in R console
    filename = "methylation.png",
    return = FALSE
)
```

```{r}
image1 <- visualise_methylation(
  #specify where to get data for locations, probabilities, and sequences
    modification_locations = methylation_info$locations,
    modification_probabilities = methylation_info$probabilities,
    sequences = methylation_info$sequences,
  #set colors and annotations
    low_colour = "white",
    high_colour = "black",
    index_annotation_lines = c(1, 23, 37),
    index_annotation_interval = 3,
    index_annotation_full_line = FALSE,
    index_annotation_always_first_base = TRUE,
    index_annotation_always_last_base = TRUE,
    other_bases_colour = "lightblue1",
    other_bases_outline_linewidth = 1,
    other_bases_outline_colour = "grey",
    modified_bases_outline_linewidth = 3,
    modified_bases_outline_colour = "black",
    margin = 0.3)

```

```{r}
image1 +
  labs(title = "test")
```


```{r}
image <- png::readPNG("test.png")
unlink("test.png")
grid::grid.newpage()
grid::grid.raster(image)
```


```{r}
image <- png::readPNG("methylation.png")
unlink("methylation.png")
grid::grid.newpage()
grid::grid.raster(image)
```

There is also a [web-based interactive version of this package](https://ejade42-ggdnavis.share.connect.posit.cloud/) for users to upload data files and customize visualizations for download, without any coding needed.


