---
title: "Demo: Sequencing Plots"
---

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}
```

```{r eval=T, echo=F, message = F, warning = F}
library(tidyverse)
library(medicaldata)
```

## ggplot extensions for specific data types

Along with statistical annotation extensions of `ggplot2`, there are also [many extension packages](https://exts.ggplot2.tidyverse.org/) designed to visualize specific types or formats of data, like sequencing data files or genomics data formats.

This section includes a brief demo of three of these packages:
* [`ggseqplot`](https://maraab23.github.io/ggseqplot/index.html)
* [`ggDNAvis`](https://ejade42.github.io/ggDNAvis/)
* [`gggenomes`](https://thackl.github.io/gggenomes/)
* [`ggtranscript`](https://dzhang32.github.io/ggtranscript/)

## Sequence plots

The [`ggseqplot`](https://maraab23.github.io/ggseqplot/articles/ggseqplot.html) package makes ggplot-style representation and summarization plots for sequence data:

| {ggseqplot} includes five types of summarization plots:
| * State distribution plots (ggseqdplot)
| * Entropy line plots (ggseqplot)
| * Modal state sequence plot (ggseqmsplot)
| * Mean time plot (ggseqmtplot)
| * Transition rate plots (ggseqtrplot)

| {ggseqplot} supports both full and subset-based representation plots, offering the following plot types:
| * Sequence index plot (ggseqiplot)
| * Sequence frequency plot (ggseqfplot)
| * Representative sequence plot (ggseqrplot)
| * Relative frequency sequence plot (ggseqrfplot)

This package is an adaptation of the R package [`TraMineR`](https://traminer.unige.ch/), which we will also install and load to access some built-in datasets.

```{r eval=FALSE}
install.packages(c("TraMineR", "ggseqplot"))
```

```{r message=FALSE}
library(TraMineR)
library(ggseqplot)
```

### Sequence strings
The [`ggDNAviz`](https://ejade42.github.io/ggDNAvis/index.html) package can make ggplot-style plots for genetic data:
| a single DNA/RNA sequence split across multiple lines,
| multiple DNA/RNA sequences, each occupying a whole line, or
| base modifications such as DNA methylation called by modified-bases models in Dorado or Guppy.

```{r eval=FALSE}
#Install package (only the first time)
install.packages("ggDNAvis")
```

```{r}
library(ggDNAvis)
library(tidyverse) #loads ggplot2 and other useful dependencies
```

This package contains a built-in dataset `example_many_sequences`, which we will use to recreate some of the plots from the official documentation for this package.

```{r}
str(example_many_sequences)
```

```{r}
head(example_many_sequences)
```

Function [visualize_many_sequences()](https://ejade42.github.io/ggDNAvis/reference/visualise_many_sequences.html) creates a ggplot-style visualization of sequences:

```{r}
# Create sequences vector using helper function extract_and_sort_sequences
sequences <- extract_and_sort_sequences(example_many_sequences)

head(sequences)
```
```{r}
#viewing in R isn't great; write to file image
visualize_many_sequences(sequences, filename = "ggDNAviz_plot.png", return = FALSE)
```

```{r}
#view created image
image <- png::readPNG("ggDNAviz_plot.png")
unlink("ggDNAviz_plot.png")
grid::grid.newpage()
grid::grid.raster(image)
```

Colors can be updating using the [sequence_colors](https://ejade42.github.io/ggDNAvis/reference/visualise_many_sequences.html#arg-sequence-colours) argument within the `visualize_many_sequences()` function.



There is also a [web-based interactive version of this package](https://ejade42-ggdnavis.share.connect.posit.cloud/) for users to upload data files and customize visualizations for download, without any coding needed.


