---
title: "Demo: Sequencing Plots"
---

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: scroll;
}

pre[class] {
  max-height: 300px;
}
```

## ggplot extensions for specific data types

Along with statistical annotation extensions of `ggplot2`, there are also [many extension packages](https://exts.ggplot2.tidyverse.org/) designed to visualize specific types or formats of data, like sequencing data files or genomics data formats.

This section includes a brief demo of [`ggDNAvis`](https://ejade42.github.io/ggDNAvis/) and [`gggenomes`](https://thackl.github.io/gggenomes/).

### Sequence strings
The [`ggDNAviz`](https://ejade42.github.io/ggDNAvis/index.html) package can make ggplot-style plots for genetic data:
> a single DNA/RNA sequence split across multiple lines,
> multiple DNA/RNA sequences, each occupying a whole line, or
> base modifications such as DNA methylation called by modified-bases models in Dorado or Guppy.

```{r eval=FALSE}
#Install package (only the first time)
install.packages("ggDNAvis")
```

```{r message=FALSE}
library(tidyverse) #loads ggplot2 and other useful dependencies
library(ggDNAvis)
```

Let's recreate some of the plots in the official package documentation, using the built-in dataset `example_many_sequences`.

First, check out the sample data:

```{r}
str(example_many_sequences)
```

```{r}
head(example_many_sequences)
```

To create a ggplot-style visualization of sequence data:

First, use one of the [data processing function](https://ejade42.github.io/ggDNAvis/reference/index.html#key-data-processing-functions) to extract and sort the sequence data from the full dataframe.

```{r}
sequences <- extract_and_sort_sequences(example_many_sequences)

head(sequences)
```
Then, use the [`visualize_many_sequqneces()` function](https://ejade42.github.io/ggDNAvis/reference/visualise_many_sequences.html). We will save the image to a file within the same function, then import it back in to view. (If following along locally, you can directly open the created png from your local file explorer to view.)

```{r}
#viewing in R isn't great; write to file image
visualize_many_sequences(sequences_vector = sequences, 
                         filename = "ggDNAviz_plot.png",
                         #don't also show image in R
                         return = FALSE)
```

```{r}
#view created image
image <- png::readPNG("ggDNAviz_plot.png")
unlink("ggDNAviz_plot.png")
grid::grid.newpage()
grid::grid.raster(image)
```

Colors for ACTG are the default ggplot-style colors, but can be updating using the [sequence_colors](https://ejade42.github.io/ggDNAvis/reference/visualise_many_sequences.html#arg-sequence-colours) argument within the `visualize_many_sequences()` function.

Note that [under the hood](https://github.com/ejade42/ggDNAvis/blob/main/R/visualise_many_sequences.R) this function makes use of ggplot2 functions. From the code for this particular function:
> result <- ggplot(image_data, aes(x = .data$x, y = .data$y)) +
>            ## Background
>            geom_tile(data = filter(image_data, .data$value == 0), width = tile_width, height = tile_height, fill = background_colour) +
>
>            ## Base boxes
>            geom_tile(data = filter(image_data, .data$value != 0), width = tile_width, height = tile_height, aes(fill = as.character(.data$value)),
>                      col = outline_colour, linewidth = outline_linewidth, linejoin = tolower(outline_join)) >+
>            scale_fill_manual(values = sequence_colours)

Note the standard ggplot syntax of `ggplot()` and use of `geom_tile()` and `scale_fill_manual()` to specify how data should be visualized. This means that while there aren't built in functions in the `ggDNAvis` package to, say, add a figure title using `labs()` as in normal ggplot images, users could in theory build their own versions of this using `geom_tile()`. But in most cases, using existing extension packages that have done the hard work already is the more appealing option.

Next, update this image to visualize methylation probabilities for these sequences.

Back to the `ggDNAvis` package, use the methylation [data processing function](https://ejade42.github.io/ggDNAvis/reference/index.html#key-data-processing-functions) to extract relevant data.

```{r}
methylation_info <- extract_and_sort_methylation(example_many_sequences)

str(methylation_info)
```
Then use the [`visualise_methylation()` function](https://ejade42.github.io/ggDNAvis/reference/visualise_methylation.html) to create a new image with this information. As before, save the image within the function and then read in to view. 

There are many customization options available:

```{r}
visualise_methylation(
  #specify where to get data for locations, probabilities, and sequences
    modification_locations = methylation_info$locations,
    modification_probabilities = methylation_info$probabilities,
    sequences = methylation_info$sequences,
  #set colors and annotations
    low_colour = "white",
    high_colour = "black",
    index_annotation_lines = c(1, 23, 37),
    index_annotation_interval = 3,
    index_annotation_full_line = FALSE,
    index_annotation_always_first_base = TRUE,
    index_annotation_always_last_base = TRUE,
    other_bases_colour = "lightblue1",
    other_bases_outline_linewidth = 1,
    other_bases_outline_colour = "grey",
    modified_bases_outline_linewidth = 3,
    modified_bases_outline_colour = "black",
    margin = 0.3,
  #save image to png file, do not also show in R console
    filename = "methylation.png",
    return = FALSE
)
```

```{r}
image <- png::readPNG("methylation.png")
unlink("methylation.png")
grid::grid.newpage()
grid::grid.raster(image)
```

There is also a [web-based interactive version of this package](https://ejade42-ggdnavis.share.connect.posit.cloud/) for users to upload data files and customize visualizations for download, without any coding needed.

## Genomic data

The package [`gggenomes`](https://thackl.github.io/gggenomes/)"extends the popular R visualization package ggplot2 by adding dedicated plot functions for genes, syntenic regions, etc. and verbs to manipulate the plot to, for example, quickly zoom in into gene neighborhoods."

```{r eval=FALSE, message=FALSE}
#Install package (only the first time)
install.packages("gggenomes")
```

```{r message=FALSE}
library(gggenomes)
```

Like most R packages, this one comes with example datasets we will use to recreate some plots from the official package documentation.

```{r}
data(package="gggenomes")
```

> emale_ava - All-versus-all whole genome alignments of 6 EMALE genomes
> emale_cogs - Clusters of orthologs of 6 EMALE proteomes
> emale_gc - Relative GC-content along 6 EMALE genomes
> emale_genes - Gene annotations if 6 EMALE genomes (endogenous virophages)
> emale_ngaros - Integrated Ngaro retrotransposons of 6 EMALE genomes
> emale_prot_ava - All-versus-all alignments 6 EMALE proteomes
> emale_seqs - Sequence index of 6 EMALE genomes (endogenous virophages)
> emale_tirs - Terminal inverted repeats of 6 EMALE genomes

The primary function is [`gggenomes()](https://thackl.github.io/gggenomes/reference/gggenomes.html) and like `ggplot2`, `gggenomes` uses the concept of "geoms" (visualization geometry) to [build up a plot](https://thackl.github.io/gggenomes/articles/gggenomes.html):

> [geom_seq()](https://thackl.github.io/gggenomes/reference/geom_seq.html) : data=seqs() for the seq track
> [geom_feat()](https://thackl.github.io/gggenomes/reference/geom_feat.html) : data=feats() for the first feat track not named “genes”
> [geom_link()](https://thackl.github.io/gggenomes/reference/geom_link.html) : data=links() for the first link track
> [geom_gene()](https://thackl.github.io/gggenomes/reference/geom_gene.html) : data=genes() for the first feat track with some extras for geneish features


Example plot:
```{r warning=FALSE}
gggenomes(
  #specify genes, sequences, links, and feature tracks
  genes = emale_genes,
  seqs = emale_seqs,
  feats = emale_tirs,
  links = emale_ava
) +
  #chromosomes
  geom_seq() + 
  #labels
  geom_bin_label() + 
  #terminal inverted repeats
  geom_feat(linewidth= 8) + 
  #genes
  geom_gene(aes(fill = strand), position = "strand") + 
  #synteny-blocks
  geom_link(offset = 0.15) 
```

This can be customized for colors and annotations mixing and matching `gggenomes` and `ggplot2` functions.

From`gggenomes`, add genomes tags.

```{r warning=FALSE}
gggenomes(
  genes = emale_genes,
  seqs = emale_seqs,
  feats = emale_tirs,
  links = emale_ava
) +
  geom_seq() + 
  geom_bin_label() + 
  geom_feat(linewidth= 8) + 
  geom_gene(aes(fill = strand), position = "strand") + 
  geom_link(offset = 0.15) +
  #add genome gene tag, format so no overlap
  geom_gene_tag(aes(label=name), 
                nudge_y=0.1, check_overlap = TRUE) 
```

From `ggplot`, change the colors using the`scale_color_discrete()` function's [palette argument](https://ggplot2.tidyverse.org/reference/scale_colour_discrete.html#arg-palette). 

There are a variety of discrete color palette options from the [`scales` package](https://scales.r-lib.org/reference/pal_brewer.html) which is installed as part of `ggplot` (which we installed as part of `tidyverse`); in this case, the [`pal_brewer()` function](https://scales.r-lib.org/reference/pal_brewer.html) (handy [cheatsheet](https://r-graph-gallery.com/38-rcolorbrewers-palettes.html) for palette names).

```{r warning=FALSE}
gggenomes(
  genes = emale_genes,
  seqs = emale_seqs,
  feats = emale_tirs,
  links = emale_ava
) +
  geom_seq() + 
  geom_bin_label() + 
  geom_feat(linewidth= 8) + 
  geom_gene(aes(fill = strand), position = "strand") + 
  geom_link(offset = 0.15) +
  geom_gene_tag(aes(label=name), 
                nudge_y=0.1, check_overlap = TRUE) +
  #change colors
  scale_fill_discrete(palette = scales::pal_brewer(palette = "Accent")) 
```

Also from `ggplot2`, adding a title using the `labs()` option.

```{r warning=FALSE}
gggenomes(
  genes = emale_genes,
  seqs = emale_seqs,
  feats = emale_tirs,
  links = emale_ava
) +
  geom_seq() + 
  geom_bin_label() + 
  geom_feat(linewidth= 8) + 
  geom_gene(aes(fill = strand), position = "strand") + 
  geom_link(offset = 0.15) +
  geom_gene_tag(aes(label=name), 
                nudge_y=0.1, check_overlap = TRUE) +
  scale_fill_discrete(palette = scales::pal_brewer(palette = "Accent")) +
  #add title
  labs(title = "gggenomes plot")
```

_Note_: If adding a `ggplot2` function does not work - either by returning an error or unexpected results - it may be that the extension package has a built-in function to do something similar. Always refer back to the package documentation (usually an online site with tutorials) to see what functions are available within-package.

And from `gggenomes`, add genomes tags.

```{r warning=FALSE}
gggenomes(
  genes = emale_genes,
  seqs = emale_seqs,
  feats = emale_tirs,
  links = emale_ava
) +
  geom_seq() + 
  geom_bin_label() + 
  geom_feat(linewidth= 8) + 
  geom_gene(aes(fill = strand), position = "strand") + 
  geom_link(offset = 0.15) +
  #add genome gene tag, format so no overlap
  geom_gene_tag(aes(label=name), 
                nudge_y=0.1, check_overlap = TRUE) +
  scale_fill_discrete(palette = scales::pal_brewer(palette = "Accent")) +
  labs(title = "gggenomes plot")
```
Since this is a ggplot object, we can save using the standard `ggsave()` options. By default, `ggsave()` will use the last plotted object as input; we could also assign the code above to an an object and specifically call that object in the code below.

```{r warning=FALSE}
ggsave("gggenomes_plot.png", width=8, height=4)
```

A [complete example](https://thackl.github.io/gggenomes/articles/emales.html) of this package, from raw data to full plot, is included the package documentation.

## Additional resources

* There are also many non-ggplot-style plotting R packages for sequence-type data ([overview in Table 1](https://pmc.ncbi.nlm.nih.gov/articles/PMC11541695/)).
* More [R color palettes](https://emilhvitfeldt.github.io/r-color-palettes/discrete.html)