---
title: "Create Statistical Plots"
---

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}
```

```{r eval=T, echo=F, message = F, warning = F}
library(tidyverse)
library(medicaldata)

smartpill <- smartpill %>% 
  mutate(Group = as.factor(Group))
```

There are many R packages that can be used to plot statistical results. This section will briefly cover `ggpmisc` and `ggpubr` for labeling ggplot plots with statistical results (model coefficients, significance) of data in that plot

## ggpmisc

The `ggpmisc` package is an extension of `ggplot2` for labeling plots with statistical results of data in the plot, for instance model equation or p-values.

```{r}
#leaving in warnings to recognize some functions are overwritten
#if concerned, can always specify exact function by first indicating package
#for instance ggplot2:heightDetail.titleGrob to use the ggplot2 version of that same name function
library(ggpmisc)
```

For our faceted scatter plot
```{r warning=FALSE}
facet_plot <- ggplot(data = smartpill, aes(x = Weight, y = SB.Time, color = Group)) +
  geom_point() +
  labs(x = "Weight (kilograms)", y = "Small bowel transit time (hours)") +
  scale_color_manual(values = c("red", "navy"),
                     breaks = c("0", "1"),
                     labels = c("Critically Ill", "Healthy"),
                     name = "Group") +
  facet_wrap(~Group, 
             labeller = as_labeller(c("0" = "Critically Ill", "1" = "Healthy"))) +
  theme_bw() +
  theme(legend.position = "none")
```

### Fit line

Fit a polynomial (by default, a basic linear model using `lm()`) using the [`stat_poly_line()` function](https://docs.r4photobiology.info/ggpmisc/reference/stat_poly_line.html).
```{r warning=FALSE}
facet_plot +
  #fits lm by default, can be customized
  #se = FALSE -> no shaded confidence interval, just line
  stat_poly_line(se = FALSE)
```

### Add function to plot

Then use the [`stat_poly_eq()` function](https://docs.r4photobiology.info/ggpmisc/reference/stat_poly_eq.html) to add the R^2^ value to the plot.
```{r warning=FALSE}
facet_plot +
  stat_poly_line(se = FALSE) +
  stat_poly_eq()
```
Further customize with full model equation and more by specifying [`use_label()`](https://docs.r4photobiology.info/ggpmisc/reference/use_label.html) within `stat_poly_eq()`.

```{r warning=FALSE}
facet_plot +
  stat_poly_line(se = FALSE) +
  #defaults include model equation and p-value
  #can be changed as needed, many options
  stat_poly_eq(use_label())
```

The [`ggpmisc` user guide](https://docs.r4photobiology.info/ggpmisc/articles/user-guide.html) contains many more options for model equations, lines, and tables.

### Save ggpmisc plot

Since this is a `gg` object, we can save it using the standard `ggsave()` function:

```{r warning=FALSE}
ggpmisc_plot <- facet_plot +
  stat_poly_line(se = FALSE) +
  stat_poly_eq(use_label())

ggsave(filename = "ggpmisc_plot.png", plot = ggpmisc_plot, height = 5, width = 7, units = "in")
```

## ggpubr

The `ggpubr` package is another extension package of `ggplot2` for adding statistical results to `ggplot` plots.

```{r message=FALSE}
library(ggpubr)
```

### Add line and function

Before, we added a trend line and model equation to our existing `ggplot` object, by adding additional functions from a new package. In the case of `ggpubr`, we can use this package's functions, which recreate ggplot-style plots with additional options.

To recreate the faceted scatterplot, use the [`ggscatter()` function, with associated add-ons](https://rpkgs.datanovia.com/ggpubr/reference/index.html#plot-two-continuous-variables) from `ggpubr`.

```{r warning=FALSE}
ggscatter(data = smartpill, x = "Weight", y = "SB.Time", facet.by = "Group",
  #color points
  color = "Group", palette = c("red", "navy"),
  #add regression line
  add = "reg.line",
  #remove confidence interval of regression line
  conf.int = FALSE,
  #add correlation coefficient
  cor.coef = TRUE
   )
```

Note that the syntax is a bit different from ggplot-style: variable names need to be in quotes, and arguments for color, etc. have different names. We'd need to do a bit more work to update labels and such. You can also do a mix-and-match approach using the `ggpubr` functions and `ggplot2` functions for the same plot.

```{r warning=FALSE}
#use ggscatter to specify data and color and regression line
ggscatter(data = smartpill, x = "Weight", y = "SB.Time",
  color = "Group", palette = c("red", "navy"),
  add = "reg.line",
  conf.int = FALSE,
  cor.coef = TRUE
   ) +
#use facet_wrap from ggplot2 to make the facets by group
facet_wrap(~Group)
#could continue with further label customization, as in prior section
```

The regression equation can be added using [`stat_regline_equation()`](https://rpkgs.datanovia.com/ggpubr/reference/stat_regline_equation.html).

```{r warning=FALSE}
ggscatter(data = smartpill, x = "Weight", y = "SB.Time",
  color = "Group", palette = c("red", "navy"),
  add = "reg.line",
  conf.int = FALSE,
  cor.coef = TRUE
   ) +
facet_wrap(~Group) +
#add regression equation
#move position so doesn't overlap with correlation values
#this may take some trial and error
stat_regline_equation(label.y = 12)
```
### Save ggpubr plot

As usual, the plot can be saved using the `ggsave()` function:

```{r}
ggpubr_plot <- ggscatter(data = smartpill, x = "Weight", y = "SB.Time",
  color = "Group", palette = c("red", "navy"),
  add = "reg.line",
  conf.int = FALSE,
  cor.coef = TRUE
   ) +
facet_wrap(~Group) +
stat_regline_equation(label.y = 12)

ggsave(filename = "ggpubr_plot.png", plot = ggpubr_plot, height = 5, width = 7, units = "in")
```

## Choosing a package

There are many more [ggplot extension](https://exts.ggplot2.tidyverse.org/index.html) packages, many of which relate to adding statistical results or context to plots. Which package you choose will depend on your visualization needs and which statistical analysis you are doing. Most standard statistical tests will be included in `ggpmisc` or `ggpubr`, but for more complex analyses, you may need to expand your package search if these options do not include it. For instance, [phylogenetic trees](https://exts.ggplot2.tidyverse.org/ggtree.html) or [comparative genomics](https://thackl.github.io/gggenomes/).

## Additional resources
* [ggpmisc website](https://docs.r4photobiology.info/ggpmisc/index.html)
* [ggpubr website](https://rpkgs.datanovia.com/ggpubr/index.html)
* Many more [ggplot extensions](https://exts.ggplot2.tidyverse.org/index.html)
* Also check out [tidyplots](https://tidyplots.org/), another good option for publication-ready plots. 