---
title: "Create Statistical Plots"
---

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}
```

```{r eval=T, echo=F, message = F, warning = F}
library(tidyverse)
library(medicaldata)
```
There are many R packages that can be used to plot statistical results. This section will briefly cover `ggpmisc` and `ggpubr` for labeling ggplot plots with statistical results (model coefficients, significance) of data in that plot

## Overview of `ggstats`

The `ggstats` package is an extension of `ggplot2` for plotting statistical results, like model coefficients and standard errors, in the style and format of ggplot objects.

```{r}
library(ggstats)
```

For example, plot the coefficents of a linear model of small bowel transit time based on weight and group.

```{r label='fix_group', echo=FALSE}
smartpill <- smartpill %>% 
  mutate(Group = as.factor(Group))
```

```{r}
#create model object using standard lm() function for linear model
mod1 <- lm(SB.Time ~ Weight + Group, data = smartpill)

#model results
summary(mod1)
```
```{r}
#use ggcoef_model() function from ggstats to plot the model coefficients
ggcoef_model(mod1)
```
P-values are indicated on the left and significance is displayed with a filled or partial circle. Coefficient estimates with standard error bars are displayed visually for each variable.

Plots can be customized further with `ggplot` options (example for [ggcoef_model](https://larmarange.github.io/ggstats/articles/ggcoef_model.html) - change title, axis labels, colors, etc.) including by adding a table of results alongside the plot.

```{r}
#output table as well as plot for model results
ggcoef_table(mod1)
```
As with `ggplot` objects, these plots can be saved using `ggsave()`.

```{r}
ggcoef_table(mod1) %>% 
  ggsave(filename = "Model1_viz.png", height = 4, width = 5, units = "in")
```


## Overview of `ggstatplot`

The `ggstatplot` package is an extension of `ggplot2` for adding statistical results, like model coefficients, directly to ggplot-style plots. 

```{r}
library(ggstatsplot)
```

### Statistics for scatterplots

Returning to the faceted scatterplot of weight vs small bowel transit time from the previous section:

```{r label='ggscatter_grouped', warning=FALSE}
facet_plot <- ggplot(data = smartpill, aes(x = Weight, y = SB.Time, color = Group)) +
  geom_point() +
  labs(x = "Weight (kilograms)", y = "Small bowel transit time (hours)") +
  scale_color_manual(values = c("red", "navy"),
                     breaks = c("0", "1"),
                     labels = c("Critically Ill", "Healthy"),
                     name = "Group") +
  facet_wrap(~Group, 
             labeller = as_labeller(c("0" = "Critically Ill", "1" = "Healthy"))) +
  theme_bw() +
  theme(legend.position = "none")

facet_plot
```

Use the [`grouped_ggscatterstats` function](https://indrajeetpatil.github.io/ggstatsplot/reference/grouped_ggscatterstats.html) to create the same visualization, but with additional plots and statistical results:

```{r}
grouped_ggscatterstats(data = smartpill, 
                       x = Weight, 
                       y = SB.Time,
                       grouping.var = Group)
```

There is a lot going on in the output plots. But, if this format is closer than not to your desired output, it is possible to change portions to suit your needs. 

Turn off statistical results, but retain plot:
```{r label='group_plain'}
grouped_ggscatterstats(data = smartpill, 
                       x = Weight, 
                       y = SB.Time,
                       grouping.var = Group,
                       #turn off subtitles
                       results.subtitle = FALSE)
```

Use standard `ggplot` options through 'ggplot.component' to further customize output:
```{r label='group_pretty'}
grouped_ggscatterstats(data = smartpill, 
                       x = Weight, 
                       y = SB.Time,
                       grouping.var = Group,
                       #turn off subtitles
                       results.subtitle = FALSE,
                       ggplot.component = labs(x = "Weight (kilograms)", y = "Small bowel transit time (hours)"))
```

Or, extract just the statistical results ([default stats used](https://indrajeetpatil.github.io/ggstatsplot/index.html?q=statistical%20test#statistical-reporting):
```{r label='get_stats'}
output2 <- grouped_ggscatterstats(data = smartpill, 
                       x = Weight, 
                       y = SB.Time,
                       grouping.var = Group)

extract_stats(output2)
```

As usual, plots can be saved with `ggsave()`.

```{r}
output2 %>% 
  ggsave(filename = "grouped_stat_viz.png", height = 5, width = 10, units = "in")
```

## Overview of `ggpmisc`

The `ggpmisc` package is an extension of `ggplot2` for labeling plots with statistical results of data in the plot, for instance model equation or p-values.

```{r}
#leaving in warnings to recognize some functions are overwritten
#if concerned, can always specify exact function by first indicating package
#for instance ggplot2:heightDetail.titleGrob to use the ggplot2 version of that same name function
library(ggpmisc)
```

For our faceted scatter plot, fit a polynomial (by default, a basic linear model using `lm()`) using the [`stat_poly_line()` function](https://docs.r4photobiology.info/ggpmisc/reference/stat_poly_line.html).
```{r warning=FALSE}
facet_plot +
  #fits lm by default, can be customized
  #se = FALSE -> no shaded confidence interval, just line
  stat_poly_line(se = FALSE)
```
Then use the [`stat_poly_eq()` function](https://docs.r4photobiology.info/ggpmisc/reference/stat_poly_eq.html) to add the R^2^ value to the plot.
```{r warning=FALSE}
facet_plot +
  stat_poly_line(se = FALSE) +
  stat_poly_eq()
```
Further customize with full model equation and more by specifying [`use_label()`](https://docs.r4photobiology.info/ggpmisc/reference/use_label.html) within `stat_poly_eq()`.

```{r warning=FALSE}
facet_plot +
  stat_poly_line(se = FALSE) +
  #defaults include model equation and p-value
  #can be changed as needed, many options
  stat_poly_eq(use_label())
```

The [`ggpmisc` user guide](https://docs.r4photobiology.info/ggpmisc/articles/user-guide.html) contains many more options for model equations, lines, and tables.

Since this is a `gg` object, we can save it using the standard `ggsave()` function:

```{r}
model_label_plot2 <- facet_plot +
  stat_poly_line(se = FALSE) +
  stat_poly_eq(use_label())

ggsave(filename = "ggpmisc_plot.png", plot = model_label_plot2, height = 5, width = 7, units = "in")
```

## Overview of `ggpubr`

The `ggpubr` package is another extension package of `ggplot2` for adding statistical results to `ggplot` plots.

```{r message=FALSE}
library(ggpubr)
```

Before, we added a trend line and model equation to our existing `ggplot` object, by adding additional functions from a new package. In the case of `ggpubr`, we can use this package's functions, which recreate ggplot-style plots with additional options.

To recreate the faceted scatterplot, use the [`ggscatter()` function, with associated add-ons](https://rpkgs.datanovia.com/ggpubr/reference/index.html#plot-two-continuous-variables) from `ggpubr`.

```{r warning=FALSE}
ggscatter(data = smartpill, x = "Weight", y = "SB.Time", facet.by = "Group",
  #color points
  color = "Group", palette = c("red", "navy"),
  #add regression line
  add = "reg.line",
  #remove confidence interval of regression line
  conf.int = FALSE,
  #add correlation coefficient
  cor.coef = TRUE
   )
```

Note that the syntax is a bit different from ggplot-style: variable names need to be in quotes, and arguments for color, etc. have different names. We'd need to do a bit more work to update labels and such. You can also do a mix-and-match approach using the `ggpubr` functions and `ggplot2` functions for the same plot.

```{r warning=FALSE}
#use ggscatter to specify data and color and regression line
ggscatter(data = smartpill, x = "Weight", y = "SB.Time",
  color = "Group", palette = c("red", "navy"),
  add = "reg.line",
  conf.int = FALSE,
  cor.coef = TRUE
   ) +
#use facet_wrap from ggplot2 to make the facets by group
facet_wrap(~Group)
#could continue with further label customization, as in prior section
```
The regression equation can be added using [`stat_regline_equation()`](https://rpkgs.datanovia.com/ggpubr/reference/stat_regline_equation.html).

```{r warning=FALSE}
ggscatter(data = smartpill, x = "Weight", y = "SB.Time",
  color = "Group", palette = c("red", "navy"),
  add = "reg.line",
  conf.int = FALSE,
  cor.coef = TRUE
   ) +
facet_wrap(~Group) +
#add regression equation
#move position so doesn't overlap with correlation values
#this may take some trial and error
stat_regline_equation(label.y = 12)
```

## Additional resources
* [ggstats website](https://larmarange.github.io/ggstats/)
* [ggstatsplot website](https://indrajeetpatil.github.io/ggstatsplot/index.html) with vignettes for included functions
* [ggpmisc website](https://docs.r4photobiology.info/ggpmisc/index.html)
* [ggpubr website](https://rpkgs.datanovia.com/ggpubr/index.html)
* Many more [ggplot extension](https://exts.ggplot2.tidyverse.org/index.html)